The very dependent recursive structure of iterated parametricity in indexed form

### 2025-12-29 paper/paper.tex
paper: prose refinements

\newcommand{\real}{\textsf{r}}

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a \emph{syntactic} way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \real A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a \emph{semantic} way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of \emph{reducibility}~\cite{Tait}, which is a semantic form of realisability, also known as \emph{glueing} in the context of category theory. In the same vein, binary parametricity over a language can be seen as defining \emph{logical relations}~\cite{Plotkin} over itself, since logical relations are a binary form of reducibility.

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form, that is, using a type-theoretic language, considering predicates of type $A \rightarrow \Prop$, or families of type $A \rightarrow \U$, or relations of type $A \times A \rightarrow \Prop$, or relevant relations -- sometimes called \emph{correspondences} -- of type $A \times A \rightarrow \U$, where $\Prop$ and $\U$ are types representing, that is classifying -- in category theory terminology --, propositions or sets, respectively. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\U. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \U$ and $\Sigma B:\U. (B \rightarrow A)$ (proof-relevant case), for $A:\U$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets, or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set. Equations specifying $\nu$-sets were given in \cite{hr25}. Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream. Such a bundle was called $\Xto{m}{n}$ while the residual stream of all $E_{n}$, $E_{n+1}$, ..., all of them precisely dependent on the bundle $E:\Xto{m}{n}$ of previous $E_0$, ..., $E_{n-1}$, was called $\Xfrom{m}{n}{D}$. The type of $E_n$, also dependent on a bundle $D:\Xto{m}{n}$, was denoted $\Xcomp{m}{n}[D=\D]$, as shown on tables~\ref{tab:coind} and~\ref{tab:core}, using an type theory as language of discourse. When $n$ is $0$, $\Xto{m}{0}$ is degenerated, that is reduced to the canonical inhabitant $\kstar$ of $\unittype$. Then, $\Xfrom{m}{0}{\kstar}$ denotes the full stream $E_0, E_1, \ldots$.

@@

Due to the intricate dependencies, the reason the whole construction is recursively well-defined is not simple. The formalisation in \cite{hr25} was recursively building a complex structure involving $3$ levels of the construction at once. In the present paper, we untangle the previous construction, making the dependencies explicit, and the new construction lends itself to a simple and complete explanation.

>>>>

From the point of view of realisability, unary parametricity over a language is a realisability using the language itself as language of realisers. Like parametricity, realisability can be expressed in a \emph{syntactic} way, that is as an interpretation from one language to another (i.e. mapping an expression $t$ of $\vdash A$ in some formal system to a derivation $t^*$ of $\vdash' |t| \textsf{r} A$ in some other formal system, for some program $|t|$ extracted from $t$ -- namely $t$ itself in the case of parametricity), or a \emph{semantic} way\footnote{This distinction between syntactic realisability and semantic realisability is itself connected to the difference between relative and absolute consistency. In the relative case, consistency of a formal system is reduced to the consistency of another formal system, and this can be done in a very weak logical system such as Primitive Recursive Arithmetic (PRA). On the other side, in the absolute case, consistency of a formal system is reduced to the consistency of the language of discourse, which, by G\"odel incompleteness theorem, if actually consistent, necessarily has to be as strong as the modelled formal system.}, that is as an interpretation in the language of discourse, which we could call in term the ``metalanguage'' (i.e. interpreting $A$ as a set $\textbf{A}$ or a relation $=_A$ of the language of discourse and interpreting a derivation $t$ of $\vdash A$ as an expression $|t|$ such that, respectively, $|t| \in \textbf{A}$ or $|t| =_A |t|$). Thus, the comparison between parametricity and realisability includes the case of \emph{reducibility}~\cite{Tait}, which is a semantic form of realisability, also known as \emph{glueing} in the context of category theory. In the same vein, binary parametricity over a language can be seen as defining \emph{logical relations}~\cite{Plotkin} over itself, since logical relations are a binary form of reducibility.

Parametricity, like realisability, reducibility, and logical relations, was most commonly handled in an \emph{indexed} form. To express it, let us use a type-theoretic language featuring a type of propositions $\Prop$, a type of sets $\U$, as well, as the standard $\Sigma$ and $\Pi$ constructions. In such language, predicates have type $A \rightarrow \Prop$, families have type $A \rightarrow \U$, relations have type $A \times A \rightarrow \Prop$, relevant relations -- sometimes called \emph{correspondences} -- have type $A \times A \rightarrow \U$. Along the indexed-fibered correspondence, that is along the equivalence between the types $A \rightarrow \Prop$ and $\Sigma B:\U. (B \stackrel{inj}{\rightarrow} A)$ (proof-irrelevant case) and $A \rightarrow \U$ and $\Sigma B:\U. (B \rightarrow A)$ (proof-relevant case), for $A:\U$, everything told in the indexed form can morally also be told in \emph{fibered} form. Such fibered approach to parametricity was for instance considered in Ghani \cite{Ghani...}, who also suggested to iterate parametricity.

@@

For $\nu$ an arity, that is $1$ in the case of predicates, families and augmented semi-simplicial sets, or $2$ in the case of relations, correspondences and semi-cubical sets, such sequence of was called a $\nu$-set. Equations specifying $\nu$-sets were given in \cite{hr25} with an accompanying formalisation in Rocq. The argument used to justify the well-foundedness of the construction was however heavy. In this paper, we present a refined well-foundedness argument which reflects the recursive essence of the construction.

\section{Description of iterated parametricity in indexed form\label{sec:reminder}}

Because of the dependency of $E_{n+1}$ into all $E_p$ for $p \leq n$, the family could not be expressed as a function. Instead, it was expressed as a stream whose $(n+1)$-th element $E_{n+1}$ was dependent on the bundle of all previous elements of the stream. Such a bundle was called $\Xto{m}{n}$ while the residual stream of all $E_{n}$, $E_{n+1}$, ..., all of them precisely dependent on the bundle $E:\Xto{m}{n}$ of previous $E_0$, ..., $E_{n-1}$, was called $\Xfrom{m}{n}{D}$. The type of $E_n$, also dependent on a bundle $D:\Xto{m}{n}$, was denoted $\Xcomp{m}{n}[D=\D]$, as shown on tables~\ref{tab:coind} and~\ref{tab:core}, using an type theory as language of discourse. When $n$ is $0$, $\Xto{m}{0}$ is degenerated, that is reduced to the canonical inhabitant $\kstar$ of $\unittype$. Then, $\Xfrom{m}{0}{\kstar}$ denotes the full stream $E_0, E_1, \ldots$.

@@
